{\boldmath Pseudoscalar top-Higgs coupling: Exploration of $\mathrm{CP}$-odd observables to resolve the sign ambiguity }

### 2016-03-05 Paper_ttbarH_Draft.tex
KK edits to Conclusions

It is important to mention that in the
$t\tbar$ rest frame the sign of the TP $\epsilon_5$ is defined through
the angle $\Delta\phi_{\ell^-\ell^+}$ [see the discussion following
  Eq.~(\ref{eq25})],
which is the angular difference between the
projections of the leptons' momenta onto the plane perpendicular to
$\vec{\tbar}$. By a similar argument to that discussed at the
beginning of Sec.~\ref{sec3.2}, we can construct an associated angular
distribution [see Eq.~(\ref{eq26})] that
will be sensitive to the sign of the pseudoscalar coupling. This
angular variable is the same as that proposed in Ref.~\cite{Ellis} as a
useful $\mathrm{CP}$-odd observable. Moreover, it is shown in
Ref.~\cite{Ellis} that the corresponding
angular distribution follows the functional
form given in Eq.~(\ref{eq27}). The ascociated shifts ($\delta$)
obtained for different values of $\kpt$ are expected to be of the same
order as those exhibited by the $\Delta\phi_1(n_t,n_{\tbar})$
distribution since this distribution
is constrained by the asymmetry $\mathcal{A}(\epsilon_5)$
[via Eq.~(\ref{eq26})],
which in turn is equal to
$\mathcal{A}(\epsilon_1)$. Also, we note that $\mathcal{A}(\epsilon_5)$ is slightly smaller 
than $\mathcal{A}(\epsilon_2)$,
as can be seen from Table \ref{table1}.
\par

@@

The
results shown in Tables \ref{table11} and \ref{table12} reveal that
with $5\times 10^3$ and $1\times 10^4$ events, respectively, the
observables $\mathcal{A}(\epsilon_4)$ and $\mathcal{A}(\epsilon_{10})$
are still useful for testing $\kpt$. Without taking into account
the loss of events that would take place in a realistic 
experimental analysis, these numbers of events
correspond to a luminosity of $\sim
300\,$-$\,600\,\mathrm{fb}^{-1}$ for the SM and even smaller for the
$\mathrm{CP}$-mixed cases (due to their larger cross section). This range
of luminosities is in principle achievable in the short term by the
LHC. We note that in order to be fully conclusive about the required
luminosity, it is important to include the effects of hadronization,
detector resolution, reconstruction efficiencies and so
forth. Such an analysis, however, is beyond the scope of this
paper.
%%%%%%%%%%%% with Cuts
%51.46/17 -0.02 +- 0.08
%45.48/17  -1.09 +- 0.07
%65.60/17   0.96 +- 0.08
%28.84/17   0.02 +- 0.07
%%%%%%%
%9.13/17     0.01 +- 0.06
%9.05/17     -0.98 +- 0.07
%19.7/17     0.80 +- 0.07
%16.59/17    0.06 +- 0.07
%
%%%%%%%%%%%% without Cuts
%16.36/17    0.0022 +-0.06
%15.25/17    -0.82 +- 0.07
%12.80/17    0.81 +- 0.07
%19.38/17    0.11 +- 0.08
%%%%%%
%7.45/17    0.005 +- 0.06
%18.17/17   -0.77 +- 0.07
%9.59/17    0.73 +- 0.06
%12.39/17   0.09 +- 0.08
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec7}
%
In this paper we have presented a collection of $\mathrm{CP}$-odd
observables based on triple product correlations that are useful for
disentangling the relative sign between the scalar ($\kp$) and a
potential pseudoscalar ($\kpt$) top-Higgs couplings in the $t\tbar H$
production with both tops decaying leptonically. We have tested the
sensitivity of the proposed observables by considering three types of
observables: asymmetries, mean values and angular distributions.\par

Through the use of spinor techniques we have written the expression
for the differential cross section of the full process in such a
manner that the production and the decay parts are separated, although
connected by the spin vectors of the top and antitop which are given
in terms of the momenta of the leptons in the final state. Moreover,
we have indentified the terms linear in $\kp$ and $\kpt$ as those
involving TPs. Among these, we have explored the three that do not
involve the momenta of the incoming quarks/gluons and at the same time
incorporate both spin vectors: $\epsilon_1\equiv \TPa$,
$\epsilon_2\equiv \TPb$ and $\epsilon_3\equiv \TPc$.\par

We have found
that $\epsilon_{1,2,3}$ allow to distinguish between the
$\mathrm{CP}$-mixed hypotheses by more than $\sim 20\sigma$ in the
case of asymmetries and $\sim 10\sigma$ in the case of mean values
when $1 \times 10^5$ simulated events are used. Furthermore, we have
shown that the angular distributions associated with these TPs are
also sensitive to the value of $\kp$ and $\kpt$ exhibiting a phase
shift that varies according to the values taken by these couplings. On
the other hand, by exploring TPs that incorporate the momenta of the
Higgs and the leptons instead of the spin vectors we can conclude that
the observables studied here appear to be more sensitive when the spin
vectors are used.\par

On the other hand, we have proposed a combination of the TPs
considered in the first place, $\epsilon_4\equiv
\epsilon_3-\epsilon_2$, that exhibits the highest sensitivity with the
separation between the $\mathrm{CP}$-mixed hypotheses being increased
by at least $3\sigma$ in the asymmetry for $1 \times 10^5$ events with
respect to $\mathcal{A}(\epsilon_{1,2,3})$. Again, when a similar
combination is constructed by using the leptons momenta instead of the
spin vectors ($\epsilon_8$), the sensitivity in the asymmetry is
decreased by $\sim 3\sigma$ compared to $\mathcal{A}(\epsilon_4)$ for
the same number of events, giving values compatible with those
obtained for the asymmetry of $\epsilon_2$ and $\epsilon_3$. \par

Taking into account the challenge of reconstructing the top and
antitop momenta due to the presence of two neutrinos in the final
state, we have proposed and tested two TP correlations that avoid this
difficulty. The first one is obtained by replacing the $t$ and $\tbar$
by the $b$ and $\bbar$ momenta ($\epsilon_9$), whereas the second
includes the visible part of the $t$ and $\tbar$ momenta
($\epsilon_{10}$). We have encountered that the latter is the most
sensitive leading to a discrimination of the $\mathrm{CP}$-mixed cases
under analysis by up to $\sim 16\sigma$.\par

Finally we have discussed the feasibility of the most sensitive
observables proposed here. We have found that with $5\times 10^3$ and
$1\times 10^4$ events respectively the observables
$\mathcal{A}(\epsilon_4)$ and $\mathcal{A}(\epsilon_{10})$ are still
useful for testing the hypotheses $(\kp=1,\kpt=\pm 1)$ giving rise to
separations of order $\sim 6\sigma$. These required number of events
would be reachable in principle in the short term by the LHC and hence
the capability of the observables studied here for testing the sign of
$\kpt/\kp$ could be probed in that context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigskip
\noindent
{\bf Acknowledgments}
\noindent 
This work has been partially supported by ANPCyT under grants No. PICT
2013-0433 and No. PICT 2013-2266, and by CONICET (NM, AS). The work of
KK was supported by the U.S.  National Science Foundation under Grant
PHY-1215785. KK also acknowledges sabbatical support from Taylor
University.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{\refname}
\let\bibsection\relax

>>>>

It is important to mention that in the
$t\tbar$ rest frame the sign of the TP $\epsilon_5$ is defined through
the angle $\Delta\phi_{\ell^-\ell^+}$ [see the discussion following
  Eq.~(\ref{eq25})],
which is the angular difference between the
projections of the leptons' momenta onto the plane perpendicular to
$\vec{\tbar}$. By a similar argument to that discussed at the
beginning of Sec.~\ref{sec3.2}, we can construct an associated angular
distribution [see Eq.~(\ref{eq26})] that
will be sensitive to the sign of the pseudoscalar coupling. This
angular variable is the same as that proposed in Ref.~\cite{Ellis} as a
useful $\mathrm{CP}$-odd observable. Moreover, it is shown in
Ref.~\cite{Ellis} that the corresponding
angular distribution follows the functional
form given in Eq.~(\ref{eq27}). The associated shifts ($\delta$)
obtained for different values of $\kpt$ are expected to be of the same
order as those exhibited by the $\Delta\phi_1(n_t,n_{\tbar})$
distribution since this distribution
is constrained by the asymmetry $\mathcal{A}(\epsilon_5)$
[via Eq.~(\ref{eq26})],
which in turn is equal to
$\mathcal{A}(\epsilon_1)$. Also, we note that $\mathcal{A}(\epsilon_5)$ is slightly smaller 
than $\mathcal{A}(\epsilon_2)$,
as can be seen from Table~\ref{table1}.
\par

@@

The
results shown in Tables \ref{table11} and \ref{table12} reveal that
with $5\times 10^3$ and $1\times 10^4$ events, respectively, the
observables $\mathcal{A}(\epsilon_4)$ and $\mathcal{A}(\epsilon_{10})$
are still useful for testing $\kpt$. Without taking into account
the loss of events that would take place in a realistic 
experimental analysis, these numbers of events
correspond to a luminosity of $\sim
300\,$-$\,600\,\mathrm{fb}^{-1}$ for the SM and even smaller for the
$\mathrm{CP}$-mixed cases (due to their larger cross section). This range
of luminosities is in principle achievable in the short term by the
LHC. We note that in order to be fully conclusive about the required
luminosity, it is important to include the effects of hadronization,
detector resolution, reconstruction efficiencies and so
forth. Such an analysis, however, is beyond the scope of this
paper.
%%%%%%%%%%%% with Cuts
%51.46/17 -0.02 +- 0.08
%45.48/17  -1.09 +- 0.07
%65.60/17   0.96 +- 0.08
%28.84/17   0.02 +- 0.07
%%%%%%%
%9.13/17     0.01 +- 0.06
%9.05/17     -0.98 +- 0.07
%19.7/17     0.80 +- 0.07
%16.59/17    0.06 +- 0.07
%
%%%%%%%%%%%% without Cuts
%16.36/17    0.0022 +-0.06
%15.25/17    -0.82 +- 0.07
%12.80/17    0.81 +- 0.07
%19.38/17    0.11 +- 0.08
%%%%%%
%7.45/17    0.005 +- 0.06
%18.17/17   -0.77 +- 0.07
%9.59/17    0.73 +- 0.06
%12.39/17   0.09 +- 0.08
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec7}
%
In this paper we have presented a collection of $\mathrm{CP}$-odd
observables based on triple product correlations
in $\ppprocess$ that are useful for
disentangling the relative sign between the scalar ($\kp$) and a
potential pseudoscalar ($\kpt$) top-Higgs couplings.
We have tested the
sensitivity of the various triple product correlations
{\bf (KK: OK? ``observables'' was repeated)}
by considering three types of
observables: asymmetries, angular distributions, and mean values.
Using these observables, we have examined several benchmark
scenarios, focusing in particular on the SM ($\kp=1$ and $\kpt=0$)
and on two ``$\mathrm{CP}$-mixed'' scenarios ($\kp=1$ and $\kpt=\pm1$).
\par

Through the use of spinor techniques we have written the expression
for the differential cross section of the full process in such a
manner that the production and the decay parts are separated, although
connected by the spin vectors of the top and antitop, which are given
in terms of the momenta of the leptons in the final state. Moreover,
we have identified the terms linear in $\kp$ and $\kpt$ as those
involving TPs. Among these, we have explored the three that do not
involve the momenta of the incoming quarks/gluons and at the same time
incorporate both spin vectors: $\epsilon_1\equiv \TPa$,
$\epsilon_2\equiv \TPb$ and $\epsilon_3\equiv \TPc$.\par

We have found
that $\epsilon_{1,2,3}$ allow one to distinguish between the
$\mathrm{CP}$-mixed scenarios by more than $\sim 20\sigma$ in the
case of asymmetries and $\sim 10\sigma$ in the case of mean values
when $1 \times 10^5$ simulated events are used. Furthermore, we have
shown that the angular distributions associated with these TPs are
also sensitive to the value of $\kp$ and $\kpt$, exhibiting a phase
shift that varies according to the values taken by these couplings.
By exploring TPs that incorporate the momenta of the
Higgs and the leptons instead of the spin vectors, we have concluded that
the observables studied here appear to be more sensitive when the spin
vectors are used.\par

We have also proposed a combination of the TPs,
$\epsilon_4\equiv \epsilon_3-\epsilon_2$,
which has a greater sensitivity than $\epsilon_1$-$\epsilon_3$.
With $1\times 10^5$ events, for example, the asymmetry associated
with this TP gives an effective separation 
between the $\mathrm{CP}$-mixed scenarios that
exceeds those coming from $\epsilon_1$-$\epsilon_3$
by at least $3\sigma$.  {\bf (KK: please check my rewording for accuracy.)}
When a similar
combination is constructed by using the leptons' momenta instead of the
spin vectors ($\epsilon_8$), the sensitivity in the asymmetry is
decreased by $\sim 3\sigma$ compared to the asymmetry associated
with $\epsilon_4$ for
the same number of events, giving values compatible with those
obtained for the asymmetries of $\epsilon_2$ and $\epsilon_3$. \par

Taking into account the challenge of reconstructing the top and
antitop momenta due to the presence of two neutrinos in the final
state, we have proposed and tested two TP correlations that avoid this
difficulty. The first one is obtained by replacing the $t$ and $\tbar$
momenta by the $b$ and $\bbar$ momenta ($\epsilon_9$), whereas the second
includes the visible part of the $t$ and $\tbar$ momenta
($\epsilon_{10}$). We have found that the latter is the more
sensitive of the two,
leading to a discrimination of the $\mathrm{CP}$-mixed cases
by up to $\sim 16\sigma$.\par

Finally, we have discussed the experimental {\bf (KK: OK?)}
feasibility of the most sensitive
observables proposed here. We have found that with $5\times 10^3$ and
$1\times 10^4$ events, respectively, the asymmetries
associated with $\epsilon_4$ and $\epsilon_{10}$ are still
useful for testing the hypotheses $(\kp=1,\kpt=\pm 1)$, giving rise to
separations of order $\sim 6\sigma$. These numbers of events
are within reach in the short term at the LHC,
so that these observables could in principle be used
to test the sign of $\kpt/\kp$ within that context.
{\bf (KK: maybe reword this sentence a bit\ldots?)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigskip
\noindent
{\bf Acknowledgments}
\noindent 
This work has been partially supported by ANPCyT under grants No. PICT
2013-0433 and No. PICT 2013-2266, and by CONICET (NM, AS). The work of
DC, EG and KK was supported by the U.S.  National Science Foundation under Grant
PHY-1215785.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{\refname}
\let\bibsection\relax
